---
title: "Hands-on Ex 1 - A Layered Grammar of Graphics: ggplot2 methods"
author: "Shreya Agarwal"
date: "17 January 2025"
date-modified: "last-modified"
execute: 
  warning: false
  message: false
  freeze: true
---

# Getting Started

## Install and launching R packages

The code chunk below uses p_load() of pacman package to check if tidyverse is packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse)
```

## Importing the data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

# Plotting different charts

```{r}
ggplot(data=exam_data)
```

```{r}
ggplot(data=exam_data,
       aes(x = MATHS))
```

## Essential Grammatical Elements in ggplot2: geom

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) + 
  geom_bar()

```

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) + 
  geom_dotplot(dotsize = 0.5)

```

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)  

```

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) + 
  geom_histogram()
```

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue")  
```

```{r}
ggplot(data = exam_data,
       aes(x = MATHS,
           fill = GENDER)) + 
  geom_histogram(bins = 20,
                 color = "grey30")
```

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()           
```

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()
```

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()            
```

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()            
```

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)        
```

## Essential Grammatical Elements in ggplot2: stat

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot()
```

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour ="red",        
               size=4)               
```

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",        
             fun="mean",           
             colour="red",          
             size=4)          
```

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)
```

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)
```

## Essential Grammatical Elements in ggplot2: Facets

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

## Essential Grammatical Elements in ggplot2: Coordinates

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)
```

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

## Essential Grammatical Elements in ggplot2: themes

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```
