---
title: "In-class Ex 7 - Visualising, Analysing and Forecasting Time-series Data: tidyverts methods"
author: "Shreya Agarwal"
date: "01 March 2025"
date-modified: "last-modified"
execute: 
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

# **7.1 Install R packages**

```{r}
pacman::p_load(tidyverse, tsibble, feasts, fable, seasonal)
```

# **7.2 Importing the data**

```{r}
ts_data <- read_csv(
  "data/visitor_arrivals_by_air.csv")
```

 `dmy()` of **lubridate** package is used to convert data type of Month-Year field from Character to Date.

```{r}
ts_data$`Month-Year` <- dmy(
  ts_data$`Month-Year`)
```

## **7.2.1 Conventional base `ts` object versus `tibble` object**

```{r}
ts_data
```

## **7.2.2 Conventional base `ts` object versus `tibble` object**

```{r}
ts_data_ts <- ts(ts_data)       
head(ts_data_ts)
```

```{r}
class(ts_data)
class(ts_data_ts)
```

## **7.2.3 Converting `tibble` object to `tsibble` object**

```{r}
ts_tsibble <- ts_data %>%
  mutate(Month = yearmonth(`Month-Year`)) %>%
  as_tsibble(index = `Month`)

ts_tsibble
```

```{r}
class(ts_tsibble)
```

# **7.3 Visualising Time-series Data**

To visualise the time-series data effectively, we need to organise the data frame from wide to long format by using [`pivot_longer()`](https://tidyr.tidyverse.org/reference/pivot_longer.html) of **tidyr** package

```{r}
ts_longer <- ts_data %>%
  pivot_longer(cols = c(2:34),
               names_to = "Country",
               values_to = "Arrivals")
```

## **7.3.1 Visualising single time-series: ggplot2 methods**

```{r}
ts_longer %>%
  filter(Country == "Vietnam") %>%
  ggplot(aes(x = `Month-Year`, 
             y = Arrivals))+
  geom_line(size = 0.5)
```

## **7.3.2 Plotting multiple time-series data with ggplot2 methods**

```{r}
ggplot(data = ts_longer, 
       aes(x = `Month-Year`, 
           y = Arrivals,
           color = Country))+
  geom_line(size = 0.5) +
  theme(legend.position = "bottom", 
        legend.box.spacing = unit(0.5, "cm"))
```

In order to provide effective comparison, [`facet_wrap()`](https://ggplot2.tidyverse.org/reference/facet_wrap.html) of **ggplot2** package is used to create small multiple line graph also known as trellis plot.

```{r}
#| fig-height: 12
ggplot(data = ts_longer, 
       aes(x = `Month-Year`, 
           y = Arrivals))+
  geom_line(size = 0.5) +
  facet_wrap(~ Country,
             ncol = 3,
             scales = "free_y") +
  theme_bw()
```

# 7.4 Visual Analysis of Time-series Data

```{r}
tsibble_longer <- ts_tsibble %>%
  pivot_longer(cols = c(2:34),
               names_to = "Country",
               values_to = "Arrivals")
```

## **7.4.1 Visual Analysis of Seasonality with Seasonal Plot**

```{r}
tsibble_longer %>%
  filter(Country == "Italy" |
         Country == "Vietnam" |
         Country == "United Kingdom" |
         Country == "Germany") %>% 
  gg_season(Arrivals)
```

## **7.4.2 Visual Analysis of Seasonality with Cycle Plot**

```{r}
tsibble_longer %>%
  filter(Country == "Vietnam" |
         Country == "Italy") %>% 
  autoplot(Arrivals) + 
  facet_grid(Country ~ ., scales = "free_y")
```

```{r}
tsibble_longer %>%
  filter(Country == "Vietnam" |
         Country == "Italy") %>% 
  gg_subseries(Arrivals)
```

# **7.5 Time series decomposition**

## **7.5.1 Single time series decomposition**

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam") %>%
  ACF(Arrivals) %>% 
  autoplot()
```

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam") %>%
  PACF(Arrivals) %>% 
  autoplot()
```

## **7.5.2 Multiple time-series decomposition**

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam" |
         `Country` == "Italy" |
         `Country` == "United Kingdom" |
         `Country` == "China") %>%
  ACF(Arrivals) %>%
  autoplot()
```

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam" |
         `Country` == "Italy" |
         `Country` == "United Kingdom" |
         `Country` == "China") %>%
  PACF(Arrivals) %>%
  autoplot()
```

## **7.5.3 Composite plot of time series decomposition**

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam") %>%
  gg_tsdisplay(Arrivals)
```

# **7.6 Visual STL Diagnostics**

## **7.6.1 Visual STL diagnostics with feasts**

```{r}
tsibble_longer %>%
  filter(`Country` == "Vietnam") %>%
  model(stl = STL(Arrivals)) %>%
  components() %>%
  autoplot()
```

# **7.7 Visual Forecasting**

## **7.7.1 Time Series Data Sampling**

```{r}
vietnam_ts <- tsibble_longer %>%
  filter(Country == "Vietnam") %>% 
  mutate(Type = if_else(
    `Month-Year` >= "2019-01-01", 
    "Hold-out", "Training"))
```

```{r}
vietnam_train <- vietnam_ts %>%
  filter(`Month-Year` < "2019-01-01")
```

## **7.7.2 Exploratory Data Analysis (EDA): Time Series Data**

```{r}
vietnam_train %>%
  model(stl = STL(Arrivals)) %>%
  components() %>%
  autoplot()
```

## **7.7.3 Fitting forecasting models**

### 7.7.3.1 Fitting Exponential Smoothing State Space (ETS) Models: fable methods

```{r}
ETS(y ~ error(c("A", "M")) 
    + trend(c("N", "A", "Ad")) 
    + season(c("N", "A", "M")))
```

### 7.7.3.2 Fitting a simple exponential smoothing (SES)

```{r}
fit_ses <- vietnam_train %>%
  model(ETS(Arrivals ~ error("A") 
            + trend("N") 
            + season("N")))
fit_ses
```

### 7.7.3.3 Examine Model Assumptions

```{r}
gg_tsresiduals(fit_ses)
```

### 7.7.3.4 The model details

```{r}
fit_ses %>%
  report()
```

### 7.7.3.5 Trend methods

```{r}
vietnam_H <- vietnam_train %>%
  model(`Holt's method` = 
          ETS(Arrivals ~ error("A") +
                trend("A") + 
                season("N")))
vietnam_H %>% report()
```

### 7.7.3.6 Damped Trend methods

```{r}
vietnam_HAd <- vietnam_train %>%
  model(`Holt's method` = 
          ETS(Arrivals ~ error("A") +
                trend("Ad") + 
                season("N")))
vietnam_HAd %>% report()
```

### 7.7.3.7 Checking for results

```{r}
gg_tsresiduals(vietnam_H)
```

```{r}
gg_tsresiduals(vietnam_HAd)
```

## **7.7.4 Fitting ETS Methods with Season: Holt-Winters**

```{r}
Vietnam_WH <- vietnam_train %>%
  model(
    Additive = ETS(Arrivals ~ error("A") 
                   + trend("A") 
                   + season("A")),
    Multiplicative = ETS(Arrivals ~ error("M") 
                         + trend("A") 
                         + season("M"))
    )

Vietnam_WH %>% report()
```

## **7.7.5 Fitting multiple ETS Models**

```{r}
fit_ETS <- vietnam_train %>%
  model(`SES` = ETS(Arrivals ~ error("A") + 
                      trend("N") + 
                      season("N")),
        `Holt`= ETS(Arrivals ~ error("A") +
                      trend("A") +
                      season("N")),
        `damped Holt` = 
          ETS(Arrivals ~ error("A") +
                trend("Ad") + 
                season("N")),
        `WH_A` = ETS(
          Arrivals ~ error("A") + 
            trend("A") + 
            season("A")),
        `WH_M` = ETS(Arrivals ~ error("M") 
                         + trend("A") 
                         + season("M"))
  )
```

## **7.7.6 The model coefficient**

```{r}
fit_ETS %>%
  tidy()
```

## **7.7.7 Step 4: Model Comparison**

```{r}
fit_ETS %>% 
  report()
```

## **7.7.8 Step 5: Forecasting future values**

```{r}
fit_ETS %>%
  forecast(h = "12 months") %>%
  autoplot(vietnam_ts, 
           level = NULL)
```

## **7.7.9 Fitting ETS Automatically**

```{r}
fit_autoETS <- vietnam_train %>%
  model(ETS(Arrivals))
fit_autoETS %>% report()
```

## **7.7.10 Fitting Fitting ETS Automatically**

```{r}
gg_tsresiduals(fit_autoETS)
```

## **7.7.11 Forecast the future values**

```{r}
fit_autoETS %>%
  forecast(h = "12 months") %>%
  autoplot(vietnam_train)
```

## **7.7.12 Visualising AutoETS model with ggplot2**

```{r}
fc_autoETS <- fit_autoETS %>%
  forecast(h = "12 months")

vietnam_ts %>%
  ggplot(aes(x=`Month`, 
             y=Arrivals)) +
  autolayer(fc_autoETS, 
            alpha = 0.6) +
  geom_line(aes(
    color = Type), 
    alpha = 0.8) + 
  geom_line(aes(
    y = .mean, 
    colour = "Forecast"), 
    data = fc_autoETS) +
  geom_line(aes(
    y = .fitted, 
    colour = "Fitted"), 
    data = augment(fit_autoETS))
```

# **7.8 AutoRegressive Integrated Moving Average(ARIMA) Methods for Time Series Forecasting: fable (tidyverts) methods**
