---
title: "Hands-on Ex 8.3 - Analytical Mapping"
author: "Shreya Agarwal"
date: "13 March 2025"
date-modified: "last-modified"
execute: 
  echo: true
  eval: true
  warning: false
  message: false
  freeze: true
---

# 8.1 Overview

## 8.1.1 Objectives

In this hands-on exercise, we will gain hands-on experience on using appropriate R methods to plot analytical maps.

## 8.1.2 Learning Outcome

By the end of this hands-on exercise, we will be able to use appropriate functions of tmap and tidyverse to perform the following tasks:

-   Importing geospatial data in rds format into R environment.
-   Creating cartographic quality choropleth maps by using appropriate tmap functions.
-   Creating rate map
-   Creating percentile map
-   Creating boxmap

# 8.2 Getting Started

## 8.2.1 Loading packages

First, we install and load **sf**, **tmap** and **tidyverse** packages into R environment.

```{r}
pacman::p_load(tmap, tidyverse, sf)
```

## 8.2.2 Importing data

For the purpose of this hands-on exercise, a prepared data set called *NGA_wp.rds* will be used. The data set is a polygon feature data.frame providing information on water point of Nigeria at the LGA level. We can find the data set in the *rds* sub-direct of the hands-on *data* folder.

```{r}
NGA_wp <- read_rds("data/rds/NGA_wp.rds")
```

# 8.3 Basic Choropleth Mapping

## 8.3.1 Visualising distribution of non-functional water point

Here, we plot a choropleth map showing the distribution of non-function water point by LGA

```{r}
p1 <- tm_shape(NGA_wp) +
  tm_fill("wp_functional",
          n = 10,
          style = "equal",
          palette = "Blues") +
  tm_borders(lwd = 0.1,
             alpha = 1) +
  tm_layout(main.title = "Distribution of functional water point by LGAs",
            legend.outside = FALSE)
```

```{r}
p2 <- tm_shape(NGA_wp) +
  tm_fill("total_wp",
          n = 10,
          style = "equal",
          palette = "Blues") +
  tm_borders(lwd = 0.1,
             alpha = 1) +
  tm_layout(main.title = "Distribution of total  water point by LGAs",
            legend.outside = FALSE)
```

```{r}
tmap_arrange(p2, p1, nrow = 1)
```

# 8.4 Choropleth Map for Rates

Our readings have emphasized the importance of mapping rates instead of raw counts, primarily because water points are not evenly distributed across space. If we fail to account for the number of water points in a given area, we risk mapping the total size of water points rather than accurately representing our intended topic.

## 8.4.1 Deriving Proportion of Functional Water Points and Non-Functional Water Points

We will tabulate the proportion of functional water points and the proportion of non-functional water points in each LGA. In the following code chunk, `mutate(`) from **dplyr** package is used to derive two fields, namely *pct_functional* and *pct_nonfunctional*.

```{r}
NGA_wp <- NGA_wp %>%
  mutate(pct_functional = wp_functional/total_wp) %>%
  mutate(pct_nonfunctional = wp_nonfunctional/total_wp)
```

## 8.4.2 Plotting map of rate

Here, we plot a choropleth map showing the distribution of percentage functional water point by LGA

```{r}
tm_shape(NGA_wp) +
  tm_fill("pct_functional",
          n = 10,
          style = "equal",
          palette = "Blues",
          legend.hist = TRUE) +
  tm_borders(lwd = 0.1,
             alpha = 1) +
  tm_layout(main.title = "Rate map of functional water point by LGAs",
            legend.outside = TRUE)
```

# 8.5 Extreme Value Maps

Extreme value maps are a specialized form of choropleth maps designed to emphasize extreme values at both ends of the scale, making it easier to identify outliers. These maps were created to incorporate spatial elements into exploratory data analysis (EDA), aligning with traditional non-spatial EDA methods (Anselin, 1994).

## 8.5.1 Percentile Map

A percentile map is a specialized type of quantile map that divides data into six predefined categories: 0-1%, 1-10%, 10-50%, 50-90%, 90-99%, and 99-100%. The breakpoints for these categories can be determined using the quantile function in base R by specifying a vector of cumulative probabilities: c(0, .01, .1, .5, .9, .99, 1). It's important to include both the starting and ending points in the calculation.

### 8.5.1.1 Data Preparation

Step 1: Exclude records with NA by using the code chunk below.

```{r}
NGA_wp <- NGA_wp %>%
  drop_na()
```

Step 2: Creating customised classification and extracting values

```{r}
percent <- c(0,.01,.1,.5,.9,.99,1)
var <- NGA_wp["pct_functional"] %>%
  st_set_geometry(NULL)
quantile(var[,1], percent)
```

Important - When extracting variables from an sf data frame, the geometry is also included. While this is useful for mapping and spatial analysis, many base R functions, such as `quantile()`, cannot process geometry data and may return errors. To resolve this, the `st_set_geometry(NULL)` function is used to remove the geometry field before performing such operations.

### 8.5.1.2 Why writing functions?

Writing a function has three big advantages over using copy-and-paste:

-   We can give a function an evocative name that makes our code easier to understand.
-   As requirements change, we only need to update code in one place, instead of many.
-   We eliminate the chance of making incidental mistakes when we copy and paste (i.e. updating a variable name in one place, but not in another).

Source: [Chapter 19: Functions](https://r4ds.had.co.nz/functions.html#functions) of **R for Data Science**.

### 8.5.1.3 Creating the get.var function

Firstly, we will write an R function as shown below to extract a variable (i.e. *wp_nonfunctional*) as a vector out of an sf data.frame.

-   arguments:
    -   vname: variable name (as character, in quotes)
    -   df: name of sf data frame
-   returns:
    -   v: vector with values (without a column name)

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% 
    st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

### 8.5.1.4 A percentile mapping function

Next, we will write a percentile mapping function by using the code chunk below.

```{r}
percentmap <- function(vnam, df, legtitle=NA, mtitle="Percentile Map"){
  percent <- c(0,.01,.1,.5,.9,.99,1)
  var <- get.var(vnam, df)
  bperc <- quantile(var, percent)
  tm_shape(df) +
  tm_polygons() +
  tm_shape(df) +
     tm_fill(vnam,
             title=legtitle,
             breaks=bperc,
             palette="Blues",
          labels=c("< 1%", "1% - 10%", "10% - 50%", "50% - 90%", "90% - 99%", "> 99%"))  +
  tm_borders() +
  tm_layout(main.title = mtitle, 
            title.position = c("right","bottom"))
}
```

### 8.5.1.5 Test drive the percentile mapping function

To run the function, type the code chunk as shown below.

```{r}
percentmap("total_wp", NGA_wp)
```

Note - This is just a bare bones implementation. Additional arguments such as the title, legend positioning just to name a few of them, could be passed to customise various features of the map.

## 8.5.2 Box Map

A box map is essentially an enhanced quartile map that includes additional lower and upper categories. If lower outliers are present, the breakpoints begin at the minimum value, with the second break at the lower fence. However, if there are no lower outliers, the first breakpoint starts at the lower fence, followed by the minimum value. In this case, no observations fall between the lower fence and the minimum value.

```{r}
ggplot(data = NGA_wp,
       aes(x = "",
           y = wp_nonfunctional)) +
  geom_boxplot()
```

-   Displaying summary statistics on a choropleth map by using the basic principles of boxplot.

-   To create a box map, a custom breaks specification will be used. However, there is a complication. The break points for the box map vary depending on whether lower or upper outliers are present.

### 8.5.2.1 Creating the boxbreaks function

The code chunk below is an R function that creating break points for a box map.

-   arguments:
    -   v: vector with observations
    -   mult: multiplier for IQR (default 1.5)
-   returns:
    -   bb: vector with 7 break points compute quartile and fences

```{r}
boxbreaks <- function(v,mult=1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  # initialize break points vector
  bb <- vector(mode="numeric",length=7)
  # logic for lower and upper fences
  if (lofence < qv[1]) {  # no lower outliers
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) { # no upper outliers
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}
```

### 8.5.2.2 Creating the get.var function

The code chunk below is an R function to extract a variable as a vector out of an sf data frame.

-   arguments:
    -   vname: variable name (as character, in quotes)
    -   df: name of sf data frame
-   returns:
    -   v: vector with values (without a column name)

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

### 8.5.2.3 Test drive the newly created function

Let’s test the newly created function

```{r}
var <- get.var("wp_nonfunctional", NGA_wp) 
boxbreaks(var)
```

### 8.5.2.4 Boxmap function

The code chunk below is an R function to create a box map. - arguments: - vnam: variable name (as character, in quotes) - df: simple features polygon layer - legtitle: legend title - mtitle: map title - mult: multiplier for IQR - returns: - a tmap-element (plots a map)

```{r}
boxmap <- function(vnam, df, 
                   legtitle=NA,
                   mtitle="Box Map",
                   mult=1.5){
  var <- get.var(vnam,df)
  bb <- boxbreaks(var)
  tm_shape(df) +
    tm_polygons() +
  tm_shape(df) +
     tm_fill(vnam,title=legtitle,
             breaks=bb,
             palette="Blues",
          labels = c("lower outlier", 
                     "< 25%", 
                     "25% - 50%", 
                     "50% - 75%",
                     "> 75%", 
                     "upper outlier"))  +
  tm_borders() +
  tm_layout(main.title = mtitle, 
            title.position = c("left",
                               "top"))
}
```

```{r}
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```
